### 2015-02-16 icfp15-servo/abstract.tex
Flesh out the outline a little.

%!TEX root = paper.tex
%
\begin{abstract}
The Servo project is an effort started at Mozilla Research to build a new web browser
engine that takes advantage of both modern parallel hardware and advances in language
technology that prevent many of the common security errors that plague all of the other
modern browsers.
The new language, Rust, enabled us to take on this project in a type-centric way instead of
the traditional C++ that relies on coding conventions, imprecise static analyses, and
fuzzers.
While many languages before have offered some combination of memory safety, parallelism,
and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
with the control over both memory and generated code that is required by systems programmers.
In this paper, \emph{systems} \emph{software} refers to programs that have some combination
of soft realtime requirements and memory, CPU, and power usage constraints that require the
programmer to be able to predict these features from their source code --- without being a
compiler developer.

Servo is currently over 300k lines of code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.
It is our hope that this work will spur additional research in new language features
suitable for a systems language that also scale up to projects the size of a web browser
engine.
\end{abstract}

>>>>

%!TEX root = paper.tex
%
\begin{abstract}
Servo is a project started at Mozilla Research to build a new web browser
engine that takes advantage of both modern parallel hardware and advances in language
technology that prevent many of the common security errors that plague all of the other
modern browsers.
The new language, Rust, enabled us to take on this project in a type-centric way instead of
the traditional C++ that relies on coding conventions, imprecise static analyses, and
fuzzers.
While many languages before have offered some combination of memory safety, parallelism,
and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
with the control over both memory and generated code that is required by systems programmers.
In this paper, \emph{systems} \emph{software} refers to programs that have some combination
of soft realtime requirements and memory, CPU, and power usage constraints that require the
programmer to be able to predict these features from their source code --- without being a
compiler developer.

Servo is currently over 300k lines of code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.
In this work, we show how a language with an affine type system and many syntactic
features from the ML family of languages can be successfully used to build state-of-the-art
systems software.
We also outline several pitfalls encountered along the way and describe some potential
areas for future research.
\end{abstract}

### 2015-02-16 icfp15-servo/intro.tex
Flesh out the outline a little.

\section{Introduction}
\label{sec:intro}

>>>>

\section{Introduction}
\label{sec:intro}
Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
was started in February of 2013.

Modern web browser engines have a similar complexity level to that of operating systems.
These engines are expected to execute arbitrary user programs within a safe sandbox, scale 
across a wide variety of hardware, manage access to shared system resources, and further 
do all of that while providing good performance and correct behavior
against sometimes-incomplete standards.
With the expectation of executing well on mobile devices and parallel hardware, the 
fundamental design assumptions have significantly changed since all of the modern
web engines were originally designed.

Servo is a new web browser engine designed to address the major environment and 
architectural changes over the last decade.
To address security, we are using a new systems programming language, Rust,
which adds static type checking to address by far the top cause of security
issues in modern web browsers --- failure in handling manual memory allocation
in \Cplusplus{}.
We scale across a wide variety of hardware by building either data- or task-parallelism, 
as appropriate, to each of the stages of the Web layout pipeline.
Finally, we are improving concurrency by reducing the simultaneous access to data
structures and using a message-passing architecture between components such as the
JavaScript engine and the rendering engine that is painting graphics to the screen.
In this experience report, we discuss the design and architecture of a modern web 
browser engine, show how modern programming language techniques --- many of which
originated in the functional programming community --- address these design 
constraints, and also touch on ongoing challenges and areas of research where we
would welcome additional community input.

### 2015-02-16 icfp15-servo/libraries.tex
Initial commit
Flesh out the outline a little.

\section{Introduction}
\label{sec:intro}

>>>>

\section{Libraries and abstractions}
\label{sec:libraries}

Many high-level languages provide abstractions over I/O, threading, parallelism, and concurrency.
This area is one where we started out with much larger aspirations in Rust but ultimately had to pare back,
for reasons very similar to the reasons that we have avoided GC by default --- it is difficult to provide a
predictable, fast implementation that works across all platforms of many systems features.
For example, thread-local storage (TLS) provides variables that are associated with a given thread.
In the presence of both lightweight (stackless) and full native threads across a variety of architectures, though,
it becomes difficult to understand what the cost of initialization and storage are for this feature.
Even lightweight threads in general, when interacting with a built-in scheduler and generic I/O system that
has to work across a variety of platforms (including Windows!) were too expensive to support in practice and
had to be removed.

Even some basic sharing, such as hash tables and vectors has proven challenging.
For example, most web browsers have implementations of vectors that allow instantiation with a default inline size,
as they have use cases where they create many thousands of vectors, nearly none of which have more than 4~elements.
In that case, removing the extra pointer indirection --- particularly if the values are of less than pointer size ---
can be a significant space savings.
But, these sorts of libraries are more highly optimized for a web browser than for a general-purpose systems language.

\subsection{Concurrency and parallelism}

Concurrency and data parallelism are two places where Rust's type system is very close to perfectly suited.
In concurrent and (safe) data parallel libraries, the data operated on by distinct threads is also itself distinct.
Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
In the case of concurrency this works quite well --- there are often individual objects that need to be owned by
a single object at a time.

TODO: code sample

Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.

TODO: code sample

### 2015-02-16 icfp15-servo/rust.tex
Initial commit
Flesh out the outline a little.

\section{Open}
\label{sec:open}

>>>>

\section{Rust}
\label{sec:rust}

TODO: basic overview

Bunch of features worked well here.

Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially. This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.

Monomorphization, as in MLton, is a great way of providing predictable output code to developers.

Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.

Biggest lesson here is by having a large project like Servo, got very early feedback on how language design decisions affected building a large systems project. Contrast with Cyclone's attempt to build something very close to C; don't actually need to be for people to be successful. Also note that many language changes had very different metrics in Servo than in the Rust compiler --- compilers do not have the same language feature usage as the likely target programs written in Rust.

\subsection{C++ interop}
Would like to have full C++ inlining.

\subsection{GC integration}
Would like to be able to plug in our own GC and stack scanning.

### 2015-02-16 icfp15-servo/safety.tex
Initial commit
Flesh out the outline a little.

\section{Rust}
\label{sec:rust}

>>>>

\section{Unsafe code}
\label{sec:safety}

There are several categories of code in Servo that cannot be written in the safe subset of Rust.

\subsection{FFI calls}

Interoperating with native C code often requires raw access to memory allocated in Rust or returns
pointers to memory that was allocated by the C libraries.
Translations to an from these raw pointers requires the use of unsafe code.

\subsection{Intrusive data structures}

Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide asymptotic
improvements in access.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.
